---
title: "Prediction Assignment for Practical Machine Learning Course"
author: "Leela"
date: "December 26, 2015"
output: html_document
---

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

##Data

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 


##What you should submit

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 


##Load Data and Libraries

First off, I'm loading all the libraries I will be using for this model.

```{r,}
library(caret)
library(rattle)
library(randomForest)
library(e1071)
```


Download the .csv files from the links above and set the working directory in the folder that contains them.
Load the .csv files in two variables, one for the training set and one for the testing set.


```{r}
set.seed(999)
training = read.csv("pml-training.csv", na.strings=c("", "NA", "NULL"))

testing = read.csv("pml-testing.csv", na.strings=c("", "NA", "NULL"))
dim(training)
dim(testing)

```


##Create Partitions in the Train Set

Here, we're partioning the Training set into two sets, one to train our model and one to test/cross validate it. I chose to include 70% of my original training set to the smaller training set, but you can choose to partition it differently.


```{r,}
inTrain <- createDataPartition(training$classe, p=0.7, list=FALSE)
train <- training[inTrain,]
test <- training[-inTrain,]
```



##Data Cleaning

Now we have to move on to cleaning the datasets. Since the NA values don't contribute anything, I chose to keep only the columns with full values. There are also some columns that contain useless (to our model) features, so I'm removing them as well.

```{r,}
train <- train[,colSums(is.na(train)) == 0]
classe <- train$classe
nums <- sapply(train, is.numeric)
train <- cbind(classe, train[,nums])
train$X <- train$num_window <-NULL
```


Now, I have to apply the same changes to the training set I partitioned before to validate my model and also to the testing set.

```{r,}
test <- test[,colSums(is.na(test)) == 0]
vclasse <- test$classe
vnums <- sapply(test, is.numeric)
test <- cbind(vclasse,test[,vnums])
colnames(test)[1] <- "classe"
test$X <- test$num_window <- NULL
```



```{r,}
testing <- testing[,colSums(is.na(testing)) == 0]
tnums <- sapply(testing, is.numeric)
testing <- testing[,tnums]
testing$X <- testing$num_window<-NULL
```


##Fitting the Model

Here, we're using Random Forests to fir our model and view the results. Please note that if you decide to run the code on your system, it might take some time.

```{r,}
fit <- train(train$classe~.,data=train, method="rf")
fit$results
```


##Estimate the Error with cross validation

Now that we have a fitted model, we are going to cross validate it with the partioned data we reserved earlier. 

Since the accuracy of the model is above 99%, I'm expecting an out of sample error of below 1%. 

```{r,}
traincontrol <- trainControl(method = "cv", number = 5)
crossvalidation <- train(test$classe~., data=test, method="rf", trControl=traincontrol)
save(crossvalidation, file="crossvalidation.RData")
load(file="./crossvalidation.RData")
crossvalidation$resample
crossvalidation$results
confusionMatrix(predict(crossvalidation, newdata=test), test$classe)
crossvalidation$finalModel
```


##Generate the prediction on the 20 cases

At last, we only have to apply the model to generate the predictions and the resulting files are to be submitted on the course website separately. 
20/20 correct answers with this one! :)

```{r,}
prediction <- predict(fit, newdata=testing)
prediction
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename, quote=FALSE, row.names=FALSE, col.names=FALSE)
  }
}
pml_write_files(prediction)
```


Thank for for reviewing my assignment. Have a great 2016 !! :)